ms_0_intro

### 2019-08-14 man/figures/resilience_metrics.Rmd
update new version

```{r load_pkg}
library("tidyverse")
library("here")
library("ggrepel")
library("grid")
library("gridExtra")
library("patchwork")
library("lemon")
library("bookdown")
```

```{r data}
mhuber <- read.csv(here("/out/anovas_resilience/huber_evi", "robust_mhuber.csv"), header = TRUE)
mhuberbai <- read.csv(here("/out/anovas_resilience/huber_bai", "robust_mhuber.csv"), header = TRUE)

@@

aux <- mhuber_filter %>% mutate(vari = '0_EVI')
aux2 <- mhuberbai_filter %>% mutate(vari = '1_BAI')
m <- bind_rows(aux2, aux)
```

```{r plot}
pd <- position_dodge(width = 0.6)

@@

rs <- m %>% filter(var == "rs") %>% 
  ggplot(aes_disturb) + themeR + 
  ylab("Resilience") + xlab('Drought event') +
    theme(plot.margin = margin(r=0, l=0, t=0, b=-2, "pt"))
# + geom_text(aes(label=Letter, y = 1.2), position=pd, show.legend = FALSE) 
```

**Figure 7.** Response *Q. pyrenaica* forests to drought in terms of resistance, recovery and resilience of greenness (EVI; left-plots) and tree growth (BAI; right-plots) for the years 2005 and 2012. For EVI we compared northern populations (*black fill circle*) with southern ones (*blue empty circle*). For BAI we compared the northern population (San Juan, SJ; *black triangle*) with the two southerns populations: Cáñar-High (CA-High; *blue empty squares*) and Cáñar-Low (CA-Low; *blue fill squares*). Different letters above error bars indicate significant *post hoc* differences between groups (see material and methods).

```{r plot_out, fig.height=5, dpi = 300, out.width = 0.8} 
grid_arrange_shared_legend(rt, rc, rs, ncol = 1, nrow = 3, position ='top')

>>>>

```{r load_pkg}
library("tidyverse")
library("here")
library("ggrepel")
library("grid")
library("gridExtra")
library("patchwork")
library("lemon")
library("bookdown")
source(here::here('/script/R/custom_functions.R'))
```

```{r data}
huber_evi <- 
  read.csv(here::here("/out/anovas_resilience/huber_evi", "robust_mhuber.csv"), header = TRUE) %>% 
  filter(var != "rrs") %>% 
  mutate(site_sorted = as.factor(
  case_when(
  site == "N" ~ "0_N",
  site == "S" ~ "1_S"
)))

huber_b_evi <- 
  read.csv(here::here("/out/anovas_resilience/huber_evi", "robust_mhuber_b.csv"), header = TRUE) %>%
  filter(var != "rrs") %>% 
  mutate(var_sorted = as.factor(
    case_when(
      var == "rt" ~ "0_rt",
      var == "rc" ~ "1_rc",
      var == "rs" ~ "2_rs",
))) %>% 
mutate(site_sorted = as.factor(
  case_when(
  site == "N" ~ "0_N",
  site == "S" ~ "1_S"
)))

huber_bai <- read.csv(here::here("/out/anovas_resilience/huber_bai", "robust_mhuber.csv"), header = TRUE) %>%
  filter(var != "rrs") %>% 
  mutate(site_sorted = as.factor(
  case_when(
  site == "SJ" ~ "0_SJ",
  site == "caH" ~ "2_CA-High",
  site == "caL" ~ "1_CA-Low"
)))

huber_b_bai <- 
  read.csv(here::here("/out/anovas_resilience/huber_bai", "robust_mhuber_b.csv"), header = TRUE) %>%
  filter(var != "rrs") %>% 
  mutate(var_sorted = as.factor(
    case_when(
      var == "rt" ~ "0_rt",
      var == "rc" ~ "1_rc",
      var == "rs" ~ "2_rs",
))) %>% 
  mutate(site_sorted = as.factor(
  case_when(
  site == "SJ" ~ "0_SJ",
  site == "caH" ~ "2_CA-High",
  site == "caL" ~ "1_CA-Low"
)))
```

```{r}
# plot_params 
pd <- position_dodge(width = 1)
aes_main <- aes(x=site_sorted, y = M.Huber, 
                ymin = lower.ci, ymax = upper.ci,
                fill = site_sorted, colour = site_sorted)
aes_aux <- aes(shape = as.factor(disturb_year), 
               label = Letter)
labelita <- as_labeller(c("0_rt" = "Resistance",
                                      "1_rc" = "Recovery", 
                                      "2_rs" = "Resilience"))

item_NS <- c("0_N" = "N", "1_S" = "S")
item_site <- c("0_SJ" = "SJ", "2_CA-High" = "CA-High", "1_CA-Low" = "CA-Low")

theme_res <- list(
  geom_errorbar(width=.1, position = pd),
  geom_point(size=7, shape=21),
  geom_text(aes(label = str_to_upper(Letter)), fontface = "bold", colour = "white"), 
  theme_bw(), xlab(""), ylab (""), 
  theme(panel.grid = element_blank(), 
        axis.text = element_text(size=9),
        legend.title = element_blank(),
        strip.background = element_rect(fill = "white"),
        strip.text = element_text(size = 11, face = "bold.italic"), 
        plot.title = element_text(size = 12, face = "bold")),
  scale_shape_manual(values = c(24,22)),
  facet_wrap(~var_sorted, scales = "free_y", labeller = labelita))

# PLOT EVI 
pevi <- ggplot(huber_b_evi, aes_main) + 
  # by disturb year 
  geom_errorbar(data=huber_evi, aes_aux, width=.1, position = pd) +
  geom_point(data=huber_evi, aes_aux, position=pd, size = 4.5, fill="white") + 
  geom_text(data = huber_evi, aes_aux, position = pd, size = 2.5) +
  scale_x_discrete(labels = item_NS) +
  scale_fill_manual(values = colores_den[3:2], labels = item_NS) +
  scale_colour_manual(values = colores_den[3:2], labels = item_NS) + theme_res +
  theme(legend.position = "top") + ggtitle ("EVI")

# PLOT BAI
pbai <- ggplot(huber_b_bai, aes_main) + 
  # by disturb year 
  geom_errorbar(data=huber_bai, aes_aux, width=.1, position = pd) +
  geom_point(data=huber_bai, aes_aux, position=pd, size = 4.5, fill="white") + 
  geom_text(data = huber_bai, aes_aux, position = pd, size = 2.5) +
  scale_x_discrete(labels = item_site) + 
  scale_fill_manual(values = rev(colores_den), labels = item_site) +
  scale_colour_manual(values = rev(colores_den), labels = item_site) + 
  theme_res + theme(legend.position = "bottom") + 
  ggtitle("BAI")

plot_res <- pevi + pbai + plot_layout(ncol=1, tag_level = "keep") + plot_annotation(tag_levels = "a") & theme(plot.tag = element_text(size = 12, face = 2)) & theme(plot.margin = margin(t=0,r=0,b=-2,l=0, "pt"))
```

**Figure 7.** Comparison of the response of  *Q. pyrenaica* forests to drought in terms of resistance, recovery and resilience of greenness (a) and tree-growth (b). For EVI we compared northern populations (*red circle*) with southern ones (*green circle*). For BAI we compared the more xeric northern population (San Juan, SJ; *red circle*) with the two southerns populations: Cáñar-High (CA-High; *blue circle*) and Cáñar-Low (CA-Low; *green circle*). Different letters indicate significant *post hoc* differences between groups (see material and methods for details).

```{r plot_out, fig.height=5, dpi = 300, out.width = 0.8} 
plot_res 
```

```{r}
# SAVE as eps 
saveImage(plot_res, path="./man/figures/svg/", dpi=300, w=22, h=25, fig= "fig_7", u = "cm") 
```

```{r, eval=FALSE, echo = FALSE}
# old plots

mhuberbai <- read.csv(here::here("/out/anovas_resilience/huber_bai", "robust_mhuber.csv"), header = TRUE)

@@

aux <- mhuber_filter %>% mutate(vari = '0_EVI')
aux2 <- mhuberbai_filter %>% mutate(vari = '1_BAI')
m <- bind_rows(aux2, aux)

pd <- position_dodge(width = 0.6)

@@

rs <- m %>% filter(var == "rs") %>% 
  ggplot(aes_disturb) + themeR + 
  ylab("Resilience") + xlab('Drought event') +
    theme(plot.margin = margin(r=0, l=0, t=0, b=-2, "pt"))
# + geom_text(aes(label=Letter, y = 1.2), position=pd, show.legend = FALSE)

# ```{r plot_out, fig.height=5, dpi = 300, out.width = 0.8} 
# grid_arrange_shared_legend(rt, rc, rs, ncol = 1, nrow = 3, position ='top')

# ```

### 2019-08-14 man/old_ms/ms.Rmd
update new version

---
title: "Relict deciduous oaks show high resilience to drought despite the effect of land-use legacies"
author: "A.J. Perez-Luque; G. Gea-Izquierdo and Zamora, R."
output:
  word_document:
    fig_caption: yes
    fig_height: 6
    fig_width: 6
    highlight: null
    reference_docx: ../templates/template_ms.docx
bibliography: ../refs/references.bib
csl: ../refs/ecology.csl
---

>>>>

---
title: "Relict deciduous oaks show high resilience to drought despite the effect of land-use legacies"
author: "A.J. Perez-Luque; G. Gea-Izquierdo and Zamora, R."
output:
  word_document:
    fig_caption: yes
    fig_height: 6
    fig_width: 6
    highlight: null
    reference_docx: ../templates/template_ms.docx
bibliography: ../refs/references.bib
csl: ../refs/ecosystems.csl
---

### 2019-08-14 man/old_ms/ms_0_intro.Rmd
update new version

The impacts of drought are also especially relevant for populations located in the rear edge of their distribution, where climatic conditions often corresponds to the species-limits and the populations are likey to be more sensitive to small variations on climate [@Hampe2005]. The response of rear-edge populations to variations on climate could be useful to estimate the adaptation of the species to predicted climatic changes [@Jump2010]. It becomes even more important for relict populations driven by historical land-uses, which can harbour high levels of intraspecific genetic diversity [@Valbuena2010]. 
Several studies analyzed the drought effects on Mediterranean tree species representing the southermost limit of their distribution [@Pasho2011; @Camarero2011; @SanchezSalguero2012; @Linares2014; @Dorado2017b; @SanchezSalguero2017; @Caminero2018], however only a few works have focused on the resilience to drought of rear-edge populations [*e.g.* @SanchezSalguero2013; @Herrero2014]. Resilience is the capacity of an ecosystem, community or individual to recover pre-disturbance structure and function after a disturbance [@Holling1973; @Lloret2011; @Hodgson2015]. The assessment of forest resilience to climatic disturbances, such as extreme droughts, provides critical information about the capacity of forests to maintain their structure and to continue providing valuable ecosystem services. This is particularly relevant for populations located in their rear-edge of their distribution, especially when they are located in Mediterranean mountains.

Remote sensing can be used for studying droughts impacts on ecosystems [@Deshayes2006; @Zhang2013; @AghaKouchak2015; @McDowell2015; @Norman2016]. Vegetation indices obtained from satellite, such us EVI (*Enhanced Vegetation Index*) are used as a proxy for assessing vegetation functioning and aboveground net primary production [@Huete2002], and changes in the patterns of these indices are considered as an indicator of the forest response to external environmental factors, such us droughts [*e.g.* @VicenteSerrano2013; @Vicca2016]. Drought impacts on vegetation using remote sensing have been analyzed at different scales [@VicenteSerrano2013; @Assal2016] and for Mediterranean area there were studies covering all region [@Gouveia2017] and others focused on local scales [@Lloret2007; @VicenteSerrano2007; @Gouveia2012; @Zribi2016].

@@

In this work we used remote sensing information and dendroecological methods to evaluate the drought impacts in canopy greenness (as a proxy to primary growth) and tree-radial growth of a Mediterranean tree species (*Quercus pyrenaica*) located in the rear-edge of their distribution.

Relict populations driven by historical land-use at their climatic (either altitudinal or latitudinal) rear-edge can be particularly vulnerable to climate change. In this study we combined remote sensing information and dendroecological methods to evaluate the drought impacts in both canopy greenness (as a proxy to primary growth) and radial growth of a Mediterranean tree species (*Quercus pyrenaica*) forests in Sierra Nevada, which represent a rear-edge of their distribution. Wealso evaluated the resilience of these forest to several extreme drought episodes. Specifically, the aims of this work were: (*i*) To quantify how two recent extreme drought events influenced primary and secondary growth of *Q. pyreancia* forests in their rear edge; (*ii*) to analyze the resilience of these forests to successive extreme drought events, both in recent times and in the long-term using time-series of radial growth; (*iii*) and to explore differences in the resilience metrics between populations located in contrasting ecological conditions within the rear edge of the distribution of this species. In addition, within the region of study we were interested to assess whether the effect of aspect and microclimatic conditions expressed in northern and southern populations of Pyrenean oak forests differ in their resistance, resilience and recovery to extreme drought events.

## 
```{r echo=FALSE, results='asis', eval=show_text}
cat("

>>>>

The impacts of drought are also especially relevant for populations located in the rear edge of their distribution, where climatic conditions often corresponds to the species-limits and the populations are likey to be more sensitive to small variations on climate [@Hampe2005]. The response of rear-edge populations to variations on climate could be useful to estimate the adaptation of the species to predicted climatic changes [@Jump2010]. It becomes even more important for relict populations driven by historical land-uses, which can harbour high levels of intraspecific genetic diversity [@Valbuena2010].

Several studies analyzed the drought effects on Mediterranean tree species representing the southermost limit of their distribution [@Pasho2011; @Camarero2011; @SanchezSalguero2012; @Linares2014; @Dorado2017b; @SanchezSalguero2017; @Caminero2018], some focused on the resilience to drought of rear-edge populations [*e.g.* @SanchezSalguero2013; @Herrero2014]. Resilience is the capacity of an ecosystem, community or individual to recover pre-disturbance structure and function after a disturbance [@Holling1973; @Lloret2011; @Hodgson2015]. The assessment of forest resilience to climatic disturbances, such as extreme droughts, provides critical information about the capacity of forests to maintain their structure and to continue providing valuable ecosystem services. This is particularly relevant for populations located in their rear-edge of their distribution, especially when they are located in Mediterranean mountains.

Here we assessed the resilience to drought of rear-edge populations of a tree Mediterranean species combining several approaches: remote sensing and tree-ring [@Jump2010].

Remote sensing can be used for studying droughts impacts on ecosystems [@Deshayes2006; @Zhang2013; @AghaKouchak2015; @McDowell2015; @Norman2016]. Vegetation indices obtained from satellite, such us EVI (*Enhanced Vegetation Index*) are used as a proxy for assessing vegetation functioning and aboveground net primary production [@Huete2002], and changes in the patterns of these indices are considered as an indicator of the forest response to external environmental factors, such us droughts [*e.g.* @VicenteSerrano2013; @Vicca2016]. Drought impacts on vegetation using remote sensing have been analyzed at different scales [@VicenteSerrano2013; @Assal2016] and for Mediterranean area there were studies covering all region [@Gouveia2017] and others focused on local scales [@Lloret2007; @VicenteSerrano2007; @Zribi2016].

@@

In this work we used remote sensing information and dendroecological methods to evaluate the drought impacts in both canopy greenness (as a proxy to primary growth) and tree-radial growth of a Mediterranean tree species (*Quercus pyrenaica* Willd.) located in Sierra Nevada, considered a rear-edge of their distribution. We are also interested in assessing resilience of these forest to several extreme drought episodes. We hypothesized that, relict populations driven by historical land-use at their climatic (either altitudinal or latitudinal) rear-edge are particularly vulnerable to climate change, and hence they will show low-values of resilience after several extreme droughts. Specifically, the aims of this work were: (*i*) To quantify how two recent extreme drought events influenced primary and secondary growth of *Q. pyreancia* forests in their rear edge; (*ii*) to analyze the resilience of these forests to successive extreme drought events, both in recent times and in the long-term using time-series of radial growth; (*iii*) and to explore differences in the resilience metrics between populations located in contrasting ecological conditions within the rear edge of the distribution of this species. In addition, within the region of study we were interested to assess whether the effect of aspect and microclimatic conditions expressed in northern and southern populations of Pyrenean oak forests differ in their resistance, resilience and recovery to extreme drought events.

```{r echo=FALSE, results='asis', eval=show_text}
cat("
